Understanding the Lomb-Scargle Periodogram

### 2017-01-16 PracticalLombScargle.tex
Finish false alarm probability section

With the dice, the calculation is easy because the number of independent
rolls is clear.
With the periodogram, though, the value at one frequency is correlated with
the value at other frequencies in a way that is quite difficult to analytically
quantify---these correlations come from the convolution with the survey
window.
Nonetheless, one common approach to estimating the distribution has been to
assume it can be modeled on some ``effective number'' of independent
frequencies $N_{eff}$, so that the FAP can be estimated as
\begin{equation}
  FAP(z) \approx 1 - \big(P_{single}(z)\big)^{N_{eff}}
  \eqlabel{FAP-neff}
\end{equation}
A very simple estimate for $N_{eff}$ is based on our arguments of the
expected peak width, $\delta f = 1/T$.
In this approximation, the number of independent peaks in a range
$0 \le f \le f_{max}$ is assumed to be $N_{eff} = f_{max} T$.
There have been various attempts to estimate this value more carefully,
both analytically and via simulations
\citep[see, e.g.][]{Horne86,Schwarzenberg-Czerny98,Cumming04,Fresura2008},
but all such approaches are necessarily only approximations.

@@

A more sophisticated treatment of the problem is that of \citet{Baluev2008},
who derived an analytic result based on the theory of stochastic processes.
The Baluev result is not an exact measure of the FAP, but rather an upper
limit that is shown to be quite close to the true FAP even
with highly structured window functons.
For the PSD-normalized periodogram, \citet{Baluev08} showed that the following
formula for the FAP provides a close upper-limit even in the case of
highly structured window functions:
\begin{equation}
  FAP(z) \approx 1 - P_{single}(z)e^{-\tau(z)}
  \eqlabel{FAP-baluev}
\end{equation}
where, for the PSD-normalized periodogram (\eq{lomb-scargle-periodogram}),
\begin{equation}
  \tau(z) = W e^{-z}\sqrt{z}
\end{equation}
and $W = f_{max} \sqrt{4\pi\ {\rm var}(t)}$ measures the effective width of the
observing window in units of the maximum frequency chosen for the analysis.
This result is not an exact measure of the FAP, but rather an approximation
to the upper limit that is valid for alias-free periodograms, {i.e.} cases
where the window function has very little structure.
Nonetheless, \citet{Baluev08} showed that this approximation holds quite well
even in the case of realistic survey windows.

\subsubsubsection{Bootstrap Method}
In the absence of a true analytic solution to the False Alarm Probability, we
can turn to computational methods such as the bootstrap.
The bootstrap method is a technique in which the statistic in question is
computed repeatedly on many random resamplings of the data in order to
approximate the distribution of that statistic
\citep[see][for a useful general discussion of this technique]{ICVG2014}.
For the periodogram, in each resampling we keep the temporal coordinates
the same, draw observations randomly with replacement from the observed
values, and then compute the maximum of the resulting periodogram.
For enough resamplings, the distribution of these maxima will approximate
the true distribution for the case with no periodic signal present.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{fig27_FAP_bootstrap}
  \caption{
    \figlabel{FAP-bootstrap}}
\end{figure}

\fig{FAP-bootstrap} compares

A nice description of the use of the bootstrap to estimate false alarm
probabilities in a periodogram is gien by \citet{Suveges12}.

he correlations are the result of the convolution with the su; nonetheless, a common approach has been to assume that the
distribution can be modeled based on some effective number of independent

This strong correlation between frequencies means that the simple formul
There have been various attempts to estimate this effective number of
frequencies and/or quantify the distribution of the maximum peak,
both analytically and via simulations
\citep[see, e.g.][]{Horne86,Schwarzenberg-Czerny98,Cumming04,
  Fresura2008,Baluev2008,Suveges12}.
Perhaps the most sophisticated approach is that of \citet{Baluev2008},
who derived an analytic result based on the theory of stochastic processes.
The Baluev result is not an exact measure of the FAP, but rather an upper
limit that is shown to be quite close to the true FAP even
with highly structured window functons.

%\begin{itemize}
%  \item Peak Width and why it's a poor proxy
%  \item False Alarm Probability and its pitfalls
%  \item Baluev method and why it's not enough
%  \item Motivation and example of Bootstrap
%  \item Is my data periodic?
%\end{itemize}

@@

(Summarize key recommendations here)

>>>>

With the dice, the calculation is easy because the number of independent
rolls is clear.
With the periodogram, though, the value at one frequency is correlated with
the value at other frequencies in a way that is quite difficult to analytically
quantify---these correlations come from the convolution with the survey
window.
Nonetheless, one common approach to estimating the distribution has been to
assume it can be modeled on some ``effective number'' of independent
frequencies $N_{eff}$, so that the FAP can be estimated as
\begin{equation}
  FAP(z) \approx 1 - \big(P_{single}(z)\big)^{N_{eff}}
  \eqlabel{FAP-neff}
\end{equation}
A very simple estimate for $N_{eff}$ is based on our arguments of the
expected peak width, $\delta f = 1/T$.
In this approximation, the number of independent peaks in a range
$0 \le f \le f_{max}$ is assumed to be $N_{eff} = f_{max} T$.
There have been various attempts to estimate this value more carefully,
both analytically and via simulations
\citep[see, e.g.][]{Horne86,Schwarzenberg-Czerny98,Cumming04,Frescura08},
but all such approaches are necessarily only approximations.

@@

A more sophisticated treatment of the problem is that of \citet{Baluev2008},
who derived an analytic result based on the theory of stochastic processes.
The Baluev result is not an exact measure of the FAP, but rather an upper
limit that is shown to be quite close to the true FAP even
with highly structured window functons.
For the PSD-normalized periodogram, \citet{Baluev2008} showed that the following
formula for the FAP provides a close upper-limit even in the case of
highly structured window functions:
\begin{equation}
  FAP(z) \approx 1 - P_{single}(z)e^{-\tau(z)}
  \eqlabel{FAP-baluev}
\end{equation}
where, for the normalized periodogram (\eq{lomb-scargle-normalized}),
\begin{equation}
  \tau(z) \approx W (1 - z)^{(N - 4)/2}\sqrt{z}
\end{equation}
and $W = f_{max} \sqrt{4\pi\ {\rm var}(t)}$ measures the effective width of the
observing window in units of the maximum frequency chosen for the analysis.
This result is not an exact measure of the FAP, but rather an approximation
to the upper limit that is valid for alias-free periodograms, {i.e.} cases
where the window function has very little structure.
Nonetheless, \citet{Baluev2008} showed that this approximation holds quite well
even in the case of realistic survey windows.

\subsubsubsection{Bootstrap Method}
In the absence of a true analytic solution to the False Alarm Probability, we
can turn to computational methods such as the bootstrap.
The bootstrap method is a technique in which the statistic in question is
computed repeatedly on many random resamplings of the data in order to
approximate the distribution of that statistic
\citep[see][for a useful general discussion of this technique]{ICVG2014}.
For the periodogram, in each resampling we keep the temporal coordinates
the same, draw observations randomly with replacement from the observed
values, and then compute the maximum of the resulting periodogram.
For enough resamplings, the distribution of these maxima will approximate
the true distribution for the case with no periodic signal present.
The bootstrap produces the most robust estimate of the FAP because it makes
no assumptions about statistical distributions, and fully accounts for survey
window effects.

Unfortunately, the computational costs of the bootstrap can be quite
prohibitive: to accurately measure small levels of FAP requires constructing
a large number of full periodograms.
If you are probing a false positive rate of $r$ among $N$ bootstrap reseamples,
you would roughly expect to find $rN \pm \sqrt{rN}$ peaks in your sample.
More concretely, for 1,000 bootstrap samples, you'll find only $\sim 10\pm 3$
peaks reflecting a 1\% false positive rate.
The random fluctuations in that count translate directly to an inability to
accurately estimate the false positive rate at that level.
A good rule-of-thumb is that to accurately characterize a FPR $r$ requires
something like $50 / r$ individual periodograms to be computed---this grows
computationally intractable quite quickly.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{fig27_FAP_bootstrap}
  \caption{Comparison of various approaches to computing the false alarm
    probability (FAP) for simulated observations with both structured and
    unstructured survey windows. Note that the approximate $N_{eff}$ method
    in its most na{\"i}ve form tends to underestimate the bootstrapped FAP,
    while the \citet{Baluev2008} method tends to overestimate the bootstrapped
    FAP, particularly for data with a highly-structured survey window.
    \figlabel{FAP-bootstrap}}
\end{figure}

\fig{FAP-bootstrap} compares these methods of estimating the false alarm
probability, for both an unstructured survey window, and a structured window
which produces the kinds of aliasing we have discussed above.
The naive approximation in this case under-estimates the FAP at nearly all
levels---so, for example, it might lead you to think a peak has a FAP of 10\%
when in fact it is closer to 30\%.
The Baluev method, by design, over-estimates the FAP, and is quite close to
the bootstraped distribution in the case of an unstructured window.
For a highly structured window, the Baluev method does not perform as well,
but still tends to over-estimate the FAP---so, for example, it might lead you
to think a peak as a FAP of 10\% when in fact it is closer to 5\%.

Finally, I will note that \citet{Suveges12} has shown the promise of a
hybrid approach of the bootstrap method and the extreme value statistics
of \citet{Baluev2008}; this has the ability to compute accurate FAPs
without the need to compute thousands of full bootstrapped periodograms.

\subsubsubsection{What Does the False Alarm Probability Measure?}

While the False Alarm Probability can be a useful concept, you should be
careful to keep in mind that it answers a very specific question:
{\it what is the probability that a signal with no periodic component would
lead to a peak of this magnitude?}
In other words, it is a measure of the false positive rate: the rate at which
we would expect to incorrectly identify a periodic signal where none
actually exists.
It tells us nothing about the false negative rate; {i.e.} the rate at which
we would expect to miss a periodic signal that is, in fact, present.
It also tells us nothing about the error rate; {i.e.} the rate at which we
would expect to identify an incorrect alias of a signal present in our data
(as we saw in \sect{failure-modes}).
Too often users are tempted to interpret the FAP too broadly, with the hope that
it would speak to these questions as well.

To the best of my knowledge, there is currently no silver bullet for answering
these broader questions within the Lomb-Scargle approach.
Perhaps the most fruitful path toward understanding of such effects for a
particular survey---with particular noise characteristics and a particular
observing window---is via simulated data.
In fact, this type of simulation is a vital component of most current and
future time-domain surveys that seek to catalog periodic objects
\citep[see, e.g.][]{opsim1, opsim2, Ivezic08LSST}
\todo{refs for SDSSII, Kepler, PanSTARRS, Gaia}.

@@

\todo{Summarize key recommendations}

\begin{itemize}
\item always use the floating-mean model---it has many benefits with virtually zero cost
\item always explore the window power first, so that you know what alias structures to expect
\item the peak of the periodogram rarely tells the entire story. Always check
  for aliases that are expected from window structure ($f = f_0 \pm n\delta f$)
  and from non-sinusoidal signals ($f = f_0 / m$).
\item avoid Bayesian methods unless you are very comfortable with the
  assumptions they make about your signal
\item avoid peak widths as a proxy for uncertainty unless you are quite
  certain that your analysis has found the correct peak.
\item avoid false alarm probabilities as a proxy for uncertainty
  unless you make very clear {\it exactly} what kind of error they measure.
  Estimate the FAP via bootstrap if computationally feasible, and otherwise
  keep in mind the failures of available estimates.
\end{itemize}

\todo{wrap up with some witty \& pithy lines of encouragement.}

\todo{acknowledge folks who helped with this.}
